---
title: "Data_Visualization"
author: "Michelle Howe"
date: "April 12, 2019"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# GGPLOT
From https://r4ds.had.co.nz/data-visualisation.html
 
## Libraries Needed
```{r}
library(tidyverse)
mtcars
```
## Scatterplot
```{r}
ggplot(data = mtcars) + #This line just says what data to use
  geom_point(mapping = aes(x = wt, y = mpg)) #This line says add points (for scatterplot)
#and specifies the x and y arguments. You always need the mapping = aes() portion.
```
## Add a third variable to Scatterplot
### Utilizing Color
```{r}
ggplot(data = mtcars) +
  geom_point(mapping = aes(x = wt, y = mpg, color = cyl))
``` 

### Utilizing size
```{r}
ggplot(data = mtcars) +
  geom_point(mapping = aes(x = wt, y = mpg, size = hp))
``` 

### Utilizing transparency
```{r}
ggplot(data = mtcars) +
  geom_point(mapping = aes(x = wt, y = mpg, alpha = hp))
``` 

### Utilizing shape
```{r}
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Petal.Length, shape = Species))
```
Shape cannot be applied to a continuous variable

## Facets
```{r}
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Petal.Length)) +
  facet_wrap(~ Species, nrow = 2) #variable used here must be discrete
```

## Smooth Geom
```{r}
ggplot(data = mtcars) +
  geom_smooth(mapping = aes(x = wt, y = mpg))
```

## Overlapping
add multiple geom functions
```{r}
ggplot(data = mtcars) +
  geom_point(mapping = aes(x = wt, y = mpg)) +
  geom_smooth(mapping = aes(x = wt, y = mpg))
```

## Bar plot
```{r}
ggplot(data = mtcars) +
  geom_bar(mapping = aes(x = cyl))
```

## Boxplots
```{r}
ggplot(data = iris, mapping = aes(x = Species, y = Petal.Length))+
  geom_boxplot()
```

Note that the mapping and aes arguments can go up in the first sentence to apply them to everything below. 
In order to flip the coordinates do the following:
```{r}
ggplot(data = iris, mapping = aes(x = Species, y = Petal.Length))+
  geom_boxplot() +
  coord_flip()
```

## Histogram
From DataCamp Introduction to the Tidyverse course
```{r}
ggplot(data = mtcars, mapping = aes(x = mpg)) +
  geom_histogram()
```

To adjust bin width, add in a binwidth argument:
```{r}
ggplot(data = mtcars, mapping = aes(x = mpg)) +
  geom_histogram(binwidth = 5)
```