---
title: "Data_Wrangling"
author: "Michelle Howe"
date: "April 12, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Examine structure of data
From https://www.datacamp.com/community/tutorials/intro-data-frame-r
```{r mtcars}
str(mtcars)
```
## Creating a data frame
From https://www.datacamp.com/community/tutorials/intro-data-frame-r
```{r}
#Define vectors
drug <- c("enalapril", "lisinopril", "losartan", "enalapril", "lisinopril", "losartan")
cost <- c(0.82, 0.43, 1.21, 0.98, 0.41, 1.10)
drugclass <- c("ACEI", "ACEI", "ARB", "ACEI", "ACEI", "ARB")
year <- c(2007, 2007, 2007, 2008, 2008, 2008)
#combine vectors into data frame using data.frame() function
HTNdrug_df <- data.frame(drug, cost, drugclass, year)
str(HTNdrug_df)
```
## Regex
From week 4 lectures and practice

### Libraries needed
From Week 4 lectures and practice
```{r}
library(stringr)
```

### Creating a regular expression pattern in R
```{r}
 #pattern <- regex("desired regex pattern" comments = TRUE)
```
### Regex Methods that can be Used (from stringr library)
```{r}
str_detect 
str_extract 
str_extract_all
str_replace
str_replace_all
str_split
```

## TIDYVERSE FUNCTIONS
From Week 5 lectures/Datacamp Introduction to Tidyverse course 

### Libraries Needed
```{r}
library(dplyr)
```
### Select
Select certain columns to look at
```{r}
HTNdrug_df %>%
  select(drug, drugclass)
HTNdrug_df %>%
  select(-drugclass)
```
Can also use ends_with and starts_with in select function

### Rename
Rename a column
This function from https://bookdown.org/rdpeng/exdata/managing-data-frames-with-the-dplyr-package.html
```{r}
HTNdrug_df %>%
  rename(therapy = drug)
```

### Filter
Look at only certain values for one or more columns
```{r}
HTNdrug_df %>%
  filter(drugclass == "ACEI")
HTNdrug_df %>%
  filter(drugclass == "ACEI", cost < 0.5)
```

### Arrange
Arrange by values in a certain column, ascending or descending values
```{r}
HTNdrug_df %>%
  arrange(cost)
HTNdrug_df %>%
  arrange(desc(cost)) #descending order
```

### Mutate
Change values in a column
```{r}
HTNdrug_df %>%
  mutate(cost = cost * 30) #change to cost per bottle of 30 tablets
HTNdrug_df %>%
  mutate(bottlecost = cost * 30) #adding column insteading changing
```
## Further TIDYVERSE FUNCTIONS
From https://r4ds.had.co.nz/tidy-data.html
 
### Libraries Needed
```{r}
library(tidyverse)
```
### Spread
Making the table wider
```{r}
newHTNdrug <- HTNdrug_df %>%
  spread(key = year, value = cost)
newHTNdrug
```

### Gather 
making table longer
```{r}
newHTNdrug %>%
  gather('2007','2008', key = "year", value = "cost")
```

### Separate
Separate a column into more than one column
```{r}
sepHTNdrug <- HTNdrug_df %>%
  separate(cost, into = c("dollars", "cents"))
sepHTNdrug
```
Separates at non-alphanumeric character by default. can add "sep =" argument  if needed

### Unite
combine columns together
```{r}
sepHTNdrug %>%
  unite(cost, dollars, cents, sep = ".")
```

### Missing Values
```{r}
HTNdrug_df %>%
  spread(key = drugclass, value = cost) %>%
  gather("ACEI", "ARB", key = "drugclass", value = "cost")
```
This left us with a mess, lets fix it
```{r}
HTNdrug_df %>%
  spread(key = drugclass, value = cost) %>%
  gather("ACEI", "ARB", key = "drugclass", value = "cost", na.rm = TRUE)
```